\pkg{depmixS4} : An \proglang{R}-package for hidden Markov models

### 2010-02-25 papers/jss/dpx4Rev.Rnw
Minor updates in jss paper
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/depmix@372 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

The central hypothesis about this data is whether indeed it is best 
described by two modes of responding rather than a single mode of 
responding with a continuous trade-off between speed and accuracy. 
The hallmark of having a discontinuity between slow versus speeded 
responding is that the switching between the two modes is assymetric
\citep[see e.g.][for a theoretical underpinning of this 
claim]{Maas1992}. The \code{fit} help page of \pkg{depmixS4} provides 
a number of examples in which the assymetricity of the switching 
process is tested; those examples are discussed at length in 
\citet{Visser2009b}.

@@

\subsubsection{Starting values}

Note also that start values for the transition parameters are provided
in this call using the \code{trstart} argument.  At this time, the
package does not provide automatic starting values.  Starting values
for parameters can be provided using three arguments: \code{instart}
for the parameters relating to the prior probabilities,
\coce{trstart}, relating the transition models, and \code{respstart}
for the parameters of the response models.  Note that the starting
values for the initial and transition models as well as of the
multinomial logit response models are interpreted as {\em probabilities},
and internally converted to multinomial logit parameters.

\subsubsection{Fitting the model and printing results}

The \code{depmix} function returns an object of class \code{depmix}
which contains the model specification, and not a fitted model. Hence,
the model needs to be fitted by calling \code{fit}:

<<>>=
fm <- fit(mod)
@
The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The \code{print} method provides summary
information on convergence, the log-likelihood and the AIC and BIC
values:
<<>>=
fm
@

@@

Applying \code{summary} to the \code{fit}ted object gives (only
transition models printed here):
<<>>=
summary(fm) 
@

@@

This provides the following fitted model parameters (only the 
response parameters are given here): 
<<>>=
summary(fm)
@

@@

The summary of the \code{fit}ted model gives the following (only the
prior model is shown here):
<<>>=
summary(fm)
@

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009}, parameters may be
fitted subject to general linear (in-)equality constraints.
Constraining and fixing parameters is done using the \code{conpat}
argument to the \code{fit}-function, which specifies for each
parameter in the model whether it's fixed (0) or free (1 or higher).
Equality constraints can be imposed by having two parameters have the
same number in the \code{conpat} vector.  When only fixed values are
required, the \code{fixed} argument can be used instead of
\code{conpat}, with zeroes for fixed parameters and other values (ones
e.g.) for non-fixed parameters.  Fitting the models subject to these
constraints is handled by the optimization routine \code{donlp2}.  To
be able to construct the \code{conpat} and/or \code{fixed} vectors one
needs the correct ordering of parameters which is briefly discussed
next before proceeding with an example.

@@

setMethod("exgaus", 
  signature(y="ANY"), 
  function(y,pstart=NULL,fixed=NULL, ...) {
    y <- matrix(y,length(y))
    x <- matrix(1) 
    parameters <- list()
    npar <- 3
    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
    if(!is.null(pstart)) {
      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
      parameters$mu <- pstart[1]
      parameters$sigma <- log(pstart[2])
      parameters$nu <- log(pstart[3])
    }
    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
    mod
  }
)
@

>>>>

The central hypothesis about this data is whether indeed it is best
described by two modes of responding rather than a single mode of
responding with a continuous trade-off between speed and accuracy.
The hallmark of having a discontinuity between slow versus speeded
responding is that the switching between the two modes is assymetric
\citep[see e.g.][for a theoretical underpinning of this
claim]{Maas1992}.  The \code{fit} help page of \pkg{depmixS4} provides
a number of examples in which the assymetricity of the switching
process is tested; those examples and other candidate models are
discussed at length in \citet{Visser2009b}.

@@

\paragraph{Starting values} Note also that start values for the
transition parameters are provided in this call using the
\code{trstart} argument.  At this time, the package does not provide
automatic starting values.  Starting values for parameters can be
provided using three arguments: \code{instart} for the parameters
relating to the prior probabilities, \coce{trstart}, relating the
transition models, and \code{respstart} for the parameters of the
response models.  Note that the starting values for the initial and
transition models as well as of the multinomial logit response models
are interpreted as {\em probabilities}, and internally converted to
multinomial logit parameters.

\paragraph{Fitting the model and printing results} The \code{depmix}
function returns an object of class \code{depmix} which contains the
model specification, and not a fitted model.  Hence, the model needs
to be fitted by calling \code{fit}:
<<>>=
fm <- fit(mod)
@

The \code{fit}-function returns an object of class
\code{depmix.fitted} which extends the \code{depmix} class, adding
convergence information (and information about constraints if these
were applied, see below).  The class has \code{print} and
\code{summary} methods to see the results.  The \code{print} method
provides summary information on convergence, the log-likelihood and
the AIC and BIC values:
<<>>=
fm 
@

@@

Applying \code{summary} to the \code{fit}ted object gives (only
transition models printed here):
<<>>=
summary(fm, which="transition") 
@

@@

This provides the following fitted model parameters (only the 
response parameters are given here): 
<<>>=
summary(fm,which="response")
@

@@

The summary of the \code{fit}ted model gives the following (only the
prior model is shown here):
<<>>=
summary(fm,which="prior")
@

%
% add more explanation about the parameters, possibly add a plot of 
% the class membership as function of age
%

@@

Using package \pkg{Rdonlp2} by \citet{Tamura2009} or \pkg{Rsolnp}
\citet{ZZZZ???}, parameters may be fitted subject to general linear
(in-)equality constraints.  Constraining and fixing parameters is done
using the \code{conpat} argument to the \code{fit}-function, which
specifies for each parameter in the model whether it's fixed (0) or
free (1 or higher).  Equality constraints can be imposed by having two
parameters have the same number in the \code{conpat} vector.  When
only fixed values are required, the \code{fixed} argument can be used
instead of \code{conpat}, with zeroes for fixed parameters and other
values (ones e.g.) for non-fixed parameters.  Fitting the models
subject to these constraints is handled by the optimization routine
\code{donlp2}.  To be able to construct the \code{conpat} and/or
\code{fixed} vectors one needs the correct ordering of parameters
which is briefly discussed next before proceeding with an example.

@@

setMethod("exgaus", 
  signature(y="ANY"), 
  function(y,pstart=NULL,fixed=NULL, ...) {
    y <- matrix(y,length(y))
    x <- matrix(1) 
    parameters <- list()
    npar <- 3
    if(is.null(fixed)) fixed <- as.logical(rep(0,npar))
    if(!is.null(pstart)) {
      if(length(pstart)!=npar) stop("length of 'pstart' must be ",npar)
      parameters$mu <- pstart[1]
      parameters$sigma <- log(pstart[2])
      parameters$nu <- log(pstart[3])
    }
    mod <- new("exgaus",parameters=parameters,fixed=fixed,x=x,y=y,npar=npar)
    mod
  }
)
